---
title: "Project 2"
output: html_notebook
---

READ ME:

```{r}
#libraries required
library(readr)
library(tree)
library(caret)
library(randomForest)
library(class)
```

Load Data

```{r}
library(readr)
df_2 = read_csv("C:/Users/Isebt/OneDrive/Documents/ISEN 413 Project 2/Project 2/Training_data_csv.csv",show_col_types = FALSE)
```

Data Exploration & pre-processing

```{r}
#After testing logistic and KNN models, I came back here to remove the 90% missing columns to see if we improve performance.

df_2 <- subset(df_2, select = -c(Q54_5_TEXT, Q55_7_TEXT))  # remove text columns

# col with more than 80% missing
high_missing_cols <- names(df_2)[colSums(is.na(df_2)) / nrow(df_2) > 0.9]

# Remove these
if(length(high_missing_cols) > 0){
  df_2 <- df_2[, !(names(df_2) %in% high_missing_cols)]
}

#replace missing with median value of column
df_2[] <- lapply(df_2, function(x){
  x[is.na(x)] <- median(x, na.rm = TRUE)
  x
})


zero_variance <- names(df_2)[sapply(df_2, function(x) sd(x) == 0)]
df_2 <- df_2[, !names(df_2) %in% zero_variance]

#We need to split before scaling for best practice
set.seed(1)
train_index_2 <- sample(1:nrow(df_2), size = 0.85 * nrow(df_2))
train_set_2 <- df_2[train_index_2, ]
test_set_2  <- df_2[-train_index_2, ]

train_predictors_2 <- subset(train_set_2, select = -Y)
train_response_2   <- train_set_2$Y

test_predictors_2 <- subset(test_set_2 , select = -Y)
test_response_2 <- test_set_2$Y

# Compute scaling parameters only from training data
train_means <- apply(train_predictors_2, 2, mean)
train_sds   <- apply(train_predictors_2, 2, sd)

# Scale training data
train_scaled_2 <- scale(train_predictors_2, center = train_means, scale = train_sds)

# Scale test data using training means/sds
test_scaled_2 <- scale(test_predictors_2, center = train_means, scale = train_sds)

# final data frames to train models
train_set_2 <- data.frame(Y = train_response_2, train_scaled_2)
test_set_2  <- data.frame(Y = test_response_2,  test_scaled_2)

```

11/27 - Trying to use rf for to identify most important variables

```{r}
library(tree)
library(randomForest)
library(caret) #package for cross validation

#resonse must be a factor so tree will be for classification
train_set_2$Y <- as.factor(train_set_2$Y)
test_set_2$Y  <- as.factor(test_set_2$Y)

# 5-fold cross validation
train_control <- trainControl(method = "cv", number = 5)


tunegrid <- expand.grid(.mtry = 2:50)  # values to try for mtry- try up to 50 at once

#perform cross validation
set.seed(1)
rf_cv <- train(Y ~ ., data = train_set_2,
               method = "rf",
               trControl = train_control,
               tuneGrid = tunegrid,
               ntree = 100)


best_idx <- which.max(rf_cv$results$Accuracy) #idx of tree with best accuracy
best_mtry <- rf_cv$results$mtry[best_idx] #mtry parameter value of tree with best accuracy

#best random forest tree
final_rf_tree <- randomForest(Y ~ ., data = train_set_2,
                              mtry = best_mtry,
                              ntree = 100,
                              importance = TRUE)


pred_prob <- predict(final_rf_tree, newdata = test_set_2, type = "prob")[, "1"]
pred_class <- ifelse(pred_prob > .5, 1, 0)
accuracy <-mean(pred_class == test_set_2$Y)



cat("Best test set accuracy (rf):", accuracy, "\n")

#Get importance as a dataframe
df_importance <- as.data.frame(importance(final_rf_tree))
df_importance$Feature <- rownames(df_importance)

# if MDA is negative, the model will perform better without that predictor
top_features <- df_importance$Feature[df_importance$MeanDecreaseAccuracy > 0] # Keep only positive MDA vals

cat("Number of top features:", length(top_features), "\n")
cat
#create train and test data using the top features only
train_top <- train_set_2[, c("Y", top_features)]
test_top  <- test_set_2[, c("Y", top_features)]
```

Lets reduce feature space even more by only considering predictors that make up 90% of importance

```{r}

df_importance <- df_importance[order(-df_importance$MeanDecreaseAccuracy), ] #sort by MDA highest to low
#utilize cumalitive sum of predictors to try and create thresholds for preidcots containing x% of importance
df_importance$cum_importance <- cumsum(df_importance$MeanDecreaseAccuracy) / sum(df_importance$MeanDecreaseAccuracy)

#reduce top features to only most important predictors that make up 90 % of the importance and power of model
top_features <- rownames(df_importance[df_importance$cum_importance <= .9, ])
length(top_features)  # check how many features remain

#create new datasets with only the top 
train_small <- train_set_2[, c("Y", top_features)]
test_small  <- test_set_2[, c("Y", top_features)]
```

Logistic Regression:

```{r}
#base Logistic regression, used to gauge if changes in data increase noise or decrease
base_model_3 <- glm(Y ~ ., data = train_small, family = binomial)

pred_2 <- predict(base_model_3, newdata = test_small, type = "response")
guess_2 <- ifelse(pred_2 > 0.50, 1, 0)

error_2 <- mean(guess_2 != test_small$Y)
accuracy_2 <- 1 - error_2

cat("Best Accuracy (baseline) :", round(accuracy_2, 4), "\n")

#LASSO

library(glmnet)

#create matrices for train and test
x2 <- model.matrix(Y ~ ., train_small)[, -1]
y2 <- as.numeric(as.character(train_small$Y))  

x2_test <- model.matrix(Y ~ ., test_small)[, -1]
y2_test <- as.numeric(as.character(test_small$Y))

# CV LASSO
set.seed(1)
cv_lasso_2 <- cv.glmnet(x2, y2, alpha = 1, family = "binomial")

plot(cv_lasso_2)
title("CV Error vs Log(Lambda)", line = 2.5)

best_lambda_lasso_2 <- cv_lasso_2$lambda.min #get lambda to make the best LASSO model

#tran the best model on all of the training data
lasso_model_2 <- glmnet(x2, y2, alpha = 1, lambda = best_lambda_lasso_2, family = "binomial")

# prediction probabilities on test set
pred_prob <- predict(lasso_model_2, newx = x2_test, type = "response")[,1]
pred_class <- ifelse(pred_prob > 0.5, 1, 0)
error_lasso <- mean(pred_class != test_small$Y)
accuracy_lasso <- 1 - error_lasso

# confusion matrix 
conf_mat <- confusionMatrix(factor(pred_class), factor(y2_test))
print(conf_mat)

# Training error 
train_pred_prob <- predict(lasso_model_2, newx = x2, type = "response")[,1]
train_pred_class <- ifelse(train_pred_prob > 0.5, 1, 0)
train_error <- mean(train_pred_class != y2)

cat("Training Error (LASSO) :", round(train_error, 4), "\n")

cat("Best test Accuracy (LASSO) :", round(accuracy_lasso, 4), "\n")
```

Best Logistic regression :0.8842

LDA & QDA

```{r}
library(caret)
#set to 10-fold cross validation
ctrl <- trainControl(method = "cv", number = 10)


#LDA
set.seed(1)
#perform cross validation for test error (no Lda parameteres)
lda_cv <- train(
  Y ~ ., 
  data = train_small,
  method = "lda",
  trControl = ctrl
)


lda_pred <- predict(lda_cv, test_small)
lda_acc <- mean(lda_pred == test_small$Y)

conf_mat <- confusionMatrix(factor(lda_pred), factor(y2_test))
print(conf_mat)
cat("LDA Test Accuracy:", round(lda_acc, 4), "\n")

#rain eror
pred_train <- predict(lda_cv, newdata = train_small)
train_accuracy <- mean(pred_train == train_small$Y)

cat("LDA Training Accuracy:", round(train_accuracy, 4), "\n")


#QDA
set.seed(1)
qda_cv <- train(
  Y ~ ., 
  data = train_small,
  method = "qda",
  trControl = ctrl
)

qda_pred <- predict(qda_cv, test_small)
qda_acc <- mean(qda_pred == test_small$Y)

cat("QDA Test Accuracy:", round(qda_acc, 4), "\n")
```

KNN

```{r}
library(class)
library(caret)

# Split features and response 
x_train_2 <- subset(train_small, select = -Y)
y_train_2 <- as.factor(train_small$Y)
x_test_2  <- subset(test_small, select = -Y)
y_test_2  <- as.factor(test_small$Y)

# 10-fold CV to tune k param , try k vals between 1-20
cv_2 <- trainControl(method = "cv", number = 10)
set.seed(1)
knn_model_2 <- train(
  x = x_train_2,
  y = y_train_2,
  method = "knn",
  tuneGrid = data.frame(k = 1:20),
  trControl = cv_2
)

# get best K
best_k <- knn_model_2$bestTune$k
print(paste("Best k:", best_k))

# Predict on test set with best knn model
predictions <- predict(knn_model_2, newdata = x_test_2)
accuracy <- mean(predictions == y_test_2)
print(paste("Test set accuracy (knn):", round(accuracy, 4)))
```

Tree Based Methods

```{r}
#Random forest retrain
library(tree)
library(randomForest)
library(caret) #package for cross validation

#resonse must be a factor so tree will be for classification
train_small$Y <- as.factor(train_small$Y)
test_small$Y  <- as.factor(test_small$Y)

#5-fold cross validation
train_control <- trainControl(method = "cv", number = 5)


tunegrid <- expand.grid(.mtry = 1:20)  # values to try for mtry- try up to 50 at once

#perform cross validation
set.seed(1)
rf_cv <- train(Y ~ ., data = train_small,
               method = "rf",
               trControl = train_control,
               tuneGrid = tunegrid,
               ntree = 100)
plot(rf_cv)



best_idx <- which.max(rf_cv$results$Accuracy) #idx of tree with best accuracy
best_mtry <- rf_cv$results$mtry[best_idx] #mtry parameter value of tree with best accuracy

#best random forest tree
final_rf_tree <- randomForest(Y ~ ., data = train_small,mtry = best_mtry,ntree = 500, ,importance = TRUE)


pred_prob <- predict(final_rf_tree, newdata = test_small, type = "prob")[, "1"]
pred_class <- ifelse(pred_prob > 0.5, 1, 0)
error_rf <- mean(pred_class != test_small$Y)
accuracy_rf <- 1 - error_rf

cat("Best test set accuracy (rf):", accuracy_rf, "\n")

#confusion marix
conf_mat <- confusionMatrix(factor(pred_class), factor(test_small$Y))
print(conf_mat)


#train error
train_pred <- predict(final_rf_tree, newdata = train_small)
train_pred_class <- ifelse(train_pred_class > 0.5, 1, 0)

train_accuracy <- mean(train_pred == train_small$Y)

cat("Training Accuracy:", round(train_accuracy, 4), "\n")
```



12/7 Testing Data
Load Test Set

```{r}
library(readr)
test_df = read_csv("C:/Users/Isebt/OneDrive/Documents/ISEN 413 Project 2/Project 2/Test_data.xlsx - Sheet1.csv",show_col_types = FALSE)

```

Preprocess the test data the same as trainin
```{r}
# Remove initial things taken out for preproccessing on train set
test_df <- subset(test_df, select = -c(Q54_5_TEXT, Q55_7_TEXT))
high_missing_cols <- names(df_2)[colSums(is.na(df_2)) / nrow(df_2) > 0.9]
if(length(high_missing_cols) > 0){
  test_df <- test_df[, !(names(test_df) %in% high_missing_cols)]
}

#Impute missing values with median the same as preprocessing
test_df[] <- lapply(test_df, function(x){
  x[is.na(x)] <- median(x, na.rm = TRUE)
  x
})

#Scale according to training mean and sd to prevent leakagee
test_predictors <- subset(test_df, select = -Y)
test_response   <- test_df$Y
numeric_features <- top_features  #restrict down to only the top features
test_scaled <- scale(test_predictors[, numeric_features],
                     center = train_means[numeric_features],
                     scale  = train_sds[numeric_features])

test_scaled <- test_scaled[, top_features]  #Only want the 70 predictors used to train the models
test_final <- data.frame(Y = test_response, test_scaled)


```

Make the best logistic regression model predict on test set
```{r}
#prep data for prediction
x2_test <- model.matrix(Y ~ ., test_final)[, -1]  
y2_test <- as.numeric(as.character(test_final$Y)) 

#Make the best Logistic model predict on test set
pred_prob <- predict(lasso_model_2, newx = x2_test, type = "response")[,1]
pred_class <- ifelse(pred_prob > 0.5, 1, 0)

#Accuracy and error
error_lasso <- mean(pred_class != y2_test)
accuracy_lasso <- 1 - error_lasso
cat("Test set error (LASSO):", error_lasso, "\n")
cat("Test set accuracy (LASSO):", accuracy_lasso, "\n")

#confusion matrix
library(caret)
conf_mat <- confusionMatrix(factor(pred_class), factor(y2_test))
print(conf_mat)



```
The test set accuracy of the best Logistic regression model W/LASSO is .88 W/ error .12

Now we will make the best random forest model predict on test data
```{r}
#best random forest tree

pred_prob <- predict(final_rf_tree, newdata = test_final, type = "prob")[, "1"]
pred_class <- ifelse(pred_prob > 0.5, 1, 0)
error_rf <- mean(pred_class != test_final$Y)
accuracy_rf <- 1 - error_rf

cat("Best random forest model error on tes set:", error_rf, "\n")
cat("Best random forest model accuracy on test set:", accuracy_rf, "\n")

#confusion marix
conf_mat <- confusionMatrix(factor(pred_class), factor(test_final$Y))
print(conf_mat)
```
The best random forest tree has .8975 accuracy!

Lastly, use LDA best model
```{r}
lda_pred <- predict(lda_cv, test_final)
lda_acc <- mean(lda_pred == test_final$Y)

cat("Best LDA model accuracy on test set:", lda_acc, "\n")
```



